Topological Equivariant Artist Model

### 2021-03-16 paper/paper.tex
D3 is explicit data structure

\section{Related Work}
The underlying structure and semantics of visualization are by definition reflected in visual representations \cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity to external concepts \cite{byrneAcquiredCodesMeaning2016}. The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, and the notion that the properties of the data and visual representation match is the basis of most evaluations of visualization. Expressiveness, as defined by Mackinlay \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987, mackinlayAutomatingDesignGraphical1986} is a measure of how much of the structure any map can encode. A fully expressive component is one that is equivariant since it has preserved all the structure in the data. 
Models of visualization evaluate this equivariance and how elements built in the model can be composed to build more complex visualizations. Mackinlay's A Presentation Tool introduced the notion of visualizations having syntax and semantics \cite{mackinlayAutomatingDesignGraphical1986} and Wilkenson described the grammar of this language \cite{wilkinsonGrammarGraphics2005}. This grammar oriented approach allows users to describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building those elements. This same limitation is explicitly stated in the functional dependency model of visualization developed by Sugibuchi\cite{sugibuchiFramwork2009} and evident in Vickers' category theory oriented framework in which semiotics are commutative. The algebraic process model by Kindlmann and Scheideggar proposes that the data and visualization transformations are commutative; while similar to our framework, it is missing any explicit mention of continuity.

On the other hand, most information visualization tools are explicitly tuned to specific continuity \cite{HeerSoftware2006,toryRethinkingVisualizationHighlevel2004}. For example, the relational database is core to libraries built on top of Grammar of Graphics\cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}. Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and  the digital humanaties oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro. Neither the table nor image model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} explicitly carry around different data representations for all the different types of visualizations they support.  Where libraries with a single core data structure have very consistent APIs, VTK and Matplotlib APIs can be rather inconsistent as every visualization has a different notion of how the data is structured.

>>>>

\section{Related Work}
The underlying structure and semantics of visualization are by definition reflected in visual representations \cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity to external concepts \cite{byrneAcquiredCodesMeaning2016}. The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, and the notion that the properties of the data and visual representation match is the basis of most evaluations of visualization. Expressiveness, as defined by Mackinlay \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987, mackinlayAutomatingDesignGraphical1986} is a measure of how much of the structure any map can encode. A fully expressive component is one that is equivariant since it has preserved all the structure in the data. 
Models of visualization evaluate this equivariance and how elements built in the model can be composed to build more complex visualizations. Mackinlay's \textit{A Presentation Tool} (APT) introduced the notion of visualizations having syntax and semantics \cite{mackinlayAutomatingDesignGraphical1986} and Wilkenson described the grammar of this language \cite{wilkinsonGrammarGraphics2005}. This grammar oriented approach allows users to describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building those elements. This same limitation is explicitly stated in the functional dependency model of visualization developed by Sugibuchi\cite{sugibuchiFramwork2009} and evident in Vickers' category theory oriented framework in which semiotics are commutative. The algebraic process model by Kindlmann and Scheideggar proposes that the data and visualization transformations are commutative; while similar to our framework, it is missing any explicit mention of continuity.

On the other hand, most information visualization tools are explicitly tuned to specific continuity \cite{HeerSoftware2006,toryRethinkingVisualizationHighlevel2004}. For example, the relational database is core to libraries built on top of Grammar of Graphics\cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}, and Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}.

Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and  the digital humanaties oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro.

Neither the table nor image model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} and D3 \cite{bostockDataDrivenDocuments2011} explicitly carry around different data representations for all the different types of visualizations they support.  Where libraries with a single core data structure have very consistent APIs, VTK and Matplotlib APIs can be rather inconsistent as every visualization has a different notion of how the data is structured.
